|

### 2023-01-20 text/ManuscriptDraft.Rmd
Submitted revised edition of manuscript

```{r eval = FALSE, echo=FALSE}
library(knitr)

knit('ManuscriptDraft.Rmd')
rmarkdown::render("ManuscriptDraft.Rmd", output_file = "ManuscriptDraft_Resubmission.docx")
system2("open","ManuscriptDraft_Resubmission.docx")
```
```{r echo=FALSE}
# Function for creating word comments in Rmarkdown
# Use: This text contains a `r word_comment("This is the comment", "comment.")`.

@@

Keywords: Arctic, *Dryas octopetala*, *Dryas integrifolia*, ecological monitoring, flowering phenology, life-history variation, tracking

@@

Word count: 
Article type: Research article

@@

 1. A fundamental challenge in pollination ecology is to understand how successful pollination is affected by the timing of flowering, the timing and frequency of pollinator visits, and the abundance and diversity of visitors. Since plant-pollinator interactions concern individual flowers, deciphering the consequences of e.g. shifts in the frequency of flower visits for plant reproductive success should ideally be based on observations of individual flowers throughout their entire flowering period. Due to the overwhelming logistical challenge of collecting such data by direct observation, it has rarely been done.

 2. Here, we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of individual flowers. Further, we present a framework for automatic tracking of flowers in time-lapse image series, evaluating tracking accuracy, and identifying unreliable tracks.

@@

The Arctic is characterized by short growing seasons and for the individual plant, timing of flowering is of critical importance. Flowering too early increases the risk of frost damage to flowers, while flowering too late limits the time for seed maturation and dispersal before the end of the growing season [@pardee2019; @ELZINGA2007; @inouye2008; @iler2021]. Further, for insect-pollinated plants, flowering must be synchronous with pollinator activity for successful reproduction. Both flower and insect phenology is partly driven by abiotic cues such as snowmelt and temperature and can thus be sensitive to climate change. Changes in floral phenology as a consequence of climate change may indirectly affect the abundance of flower-dependent insect species [@hoye2013] and these effects can be stronger than the direct effects of climate change [@ogilvie2017]. However, the relation between floral phenology and reproductive success is not well understood. For example, @bolmgren2015 found that for an insect-pollinated plant species, early flowering was associated with increased fruit-set but late flowering was associated with higher seed mass. If we are to understand the intricate dynamics of plant-pollinator interactions, tools that aid efficient and comprehensive data collection are essential.

@@

Cameras have become an ubiquitous tool for non-invasive and efficient monitoring of natural processes [@howe2017; @glover2019], but the implementation of image-based data collection has outpaced the development of methods for processing and analysing the big data output [@weinstein2018; @farley2018]. Gradually, computer vision and deep learning tools are being developed and applied for automatic extraction of relevant data from images [@willi2019; @tabak2019; @hoye2021]. Indeed, automatic image-based monitoring of flowering phenology coupled with deep learning has facilitated collection of phenology data for specific species at very high temporal resolution [@mann2022]. While automated detection and counting of flowers in image-series, as demonstrated in @mann2022, can produce detailed information about floral biology, phenological responses at the level of individual flowers may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be directly explored on the basis of individual level floral phenology data and such information is critical for investigating the association between reproductive success and timing of flowering and/or flower longevity. Individual level data would also aid the understanding of whether flower visitation rates and/or seed set depends on the timing of flowering for the individual flower. Thus, automated recording of floral phenology by tracking the presence of individual flowers in image-series of permanent plots would be ideal and the feasibility of this as a step subsequent of flower detection was explored in this study.

@@

Most multiple object tracking (MOT) methods focus on tracking of humans [@luo2021; @milan2016], and are not necessarily directly transferable to other use cases. Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], rely on Kalmann filtering to predict the future location of an object based on its previous velocity [@kalmann]. However, Kalman filtering assumes linear object motion and does not handle abrupt movement well. Flowers are constrained in movement by their stalk and violate this assumption. Other tracking approaches, such as DeepSORT [@deepsort2017], use tracking-by-detection in which convolutional neural networks (CNNs) are applied to continuously distinguish individuals through time, thus aiding the tracking algorithm. Individual flowers, however, are unlikely to be clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame, which prevents this approach. Finally, some algorithms require manual initiation of a fixed number of tracks in the first frame and do not allow objects to enter or disappear during tracking [@hu2012]. In full-season monitoring data, flowers should continuously be tracked as they bolt and wilt throughout the season Additionally, a flower tracking method must handle flowers disappearing and appearing at the edges of the frame. Clearly, individual flower tracking is a challenging task.

@@

In MOT, two fundamental errors can occur. First, a single trajectory may be split into several tracks, an error known as fragmentation. This can happen when an object is temporarily occluded or have disappeared from view and a new track is erroneously initiated when it reappears, or when identity shifts occur between tracks, for example when they switch positions between frames. Fragmentation is especially prone in settings with high densities of objects. Second, merging occurs when several trajectories are assigned the same track identity. This can occur when a track is not correctly terminated at the end of a trajectory but rather continues on the trajectory of another object. Here, importantly, the two individual trajectories can be tracked without internal errors (i.e. without mismatches).

@@

![Figure 2: Simulated flower detection data to demonstrate the tracking algorithm and the effect of the three tracking parameters, max distance, max gap, and running mean (refer to Automatic flower tracking section for definitions), as well as the significance of frame rate for reliable tracking. A: 10 points per flower, rF (radius of circle delimiting flower points) = 200, rM (radius of circles delimiting distance between consecutive points) = 100. Arrow shows the last point of flower a and the first point of flower b. B: 10 points per flower, rF = 200, rM = 100. C: 500 points per flower, rF = 200, rM = 200. D: 10,000 points per flower, rF = 100, rM = 1.](../figures/simulatedData.png){width="30%"}

@@

# 
![Table 1: Number of images and number of individual flowers for each of the five time-lapse image series.](../figures/CameraOverview.png){width="40%"}

@@

![Figure 3: Examples of annotated images for 2019 NARS C. A: Onset of flowering at DOY 145. B: Peak flowering at day of year (DOY) 151 (52 annotated flowers). C: 6 days after peak flowering (DOY 157). Flower bounding boxes are colour-coded by flower ID. D: Centroids of all flower annotations in the series, coloured by flower ID, showing how flowers may overlap in space.](../figures/AnnotationExamples_v2_4.png){width="20%"}

@@

We present a framework for tracking, filtering, and evaluating tracking of objects in time-lapse image series. The tracking algorithm (suppl. 1) is based on centroid tracking [@rahman2020; @bjerge2021]. That is, objects with minimum displacement between frames are associated. Centroids are derived from bounding boxes around objects stemming from manual annotation or automatic detection of objects. Thus, the input data for the tracking algorithm are bounding box coordinates (top left corner and bottom right corner) for each object [@everingham2010]. The algorithm has a set of user-adjusted parameters that can optimize tracking accuracy. The parameters are particularly relevant for tracking of objects that are constrained to a specific area such as flowers.

Several issues complicate the task of tracking individual flowers through time-lapse image series. First, as flowers move, they may temporarily occlude the view of other flowers and flowers close to the edge of the frame may move in and out of view between images. This can cause a track to be lost and fragmentation to occur (Fig. 4A). However, if a flower reappears shortly after having disappeared from view, it is a reasonable assumption that it is the same individual. The parameter **max gap** sets the number of frames a flower can be absent before reappearing without triggering track termination and initiation of a new track. Concurrently, the parameter can prevent track termination owing to false negatives.

Second, a standard centroid tracking algorithm will associate a point with a track based on minimum distance only, disregarding the absolute distance between centroids, which may lead to tracking errors (Fig. 4B). However, as flowers are constrained in their displacement, we can assume a maximum distance between points and force the initiation of tracks for points that exceed this threshold. We set this threshold with the parameter **max distance,** measured as the euclidean distance in number of pixels between objects in neighboring frames.

Third, reliable tracking based on distance requires that movement of individuals are captured at sufficient frequency to prevent identity shifts. Stand-alone image-based monitoring of the relatively slow process of flower development through full growing seasons (several months) is likely to be done with low frame rate, limited by power consumption and storage capacity both on and off the camera. Here, we used several images per day and while this temporal resolution captures the phenology of the flowers at very high detail, they fail to capture the full movement of flowers. For example, as the wind shifts, the flower heads change direction, which in the image series happens instantaneously (i.e., between two consecutive frames) (Fig. 4C). The fact that the frame rate is insufficient to capture smooth movement means that bounding boxes around a single individual in two consecutive frames cannot be associated based on overlap (e.g. intersection over union) as there often is no overlap. With a high number of flowers in close vicinity to each other, there is a substantial risk of identity shifts between flowers (Fig. 1C;4C). Additionally, as the flowers are constrained by their stalk, there is a limit to the distance they can move, which prevents prediction of future location of an individual based on its previous trajectory. Here, we use the tracking parameter **running mean**, with which we can base tracking on the distance between a point in the current frame and the running mean of the positions of a number of previous points in a track, taking advantage of the fact that a flower moves around a central point. Increasing the value for this parameter decreases the weighting of movement between individual frames. Setting the parameter to 1 means that tracking is based on the distance to the centroid in the previous frame only.

![Figure 4: Centroid tracking may produce erroneous associations when objects move or disappear between frames. A: A track may be lost when a flower temporarily disappears from view. If gaps in tracks are not allowed during tracking, a new track will erroneously be initiated when the flower reappears. B: A flower may be linked to an incorrect track despite large distance between them. Here, a flower disappears from view and another reappears between two frames. If no threshold for maximum distance between an object and a track is set, the reappearing flower will be linked to the track of the disappearing flower based on shortest distance despite the absolute distance being large. C: A flower may be linked to an incorrect track despite shortest distance to the running mean of the correct track. As flowers move around a central point, it is relevant to base tracking on a running mean of previous positions. Dots show position of flower in previous frame. Crosses show running mean of positions in x \> 1 number of previous frames. Flowers, dots and crosses coloured according to identity.](../figures/errorFigure.png){width="100%"}

@@

## Flowers
Manual tracking of individual flowers derived from image-based plot monitoring enabled detailed information on floral phenology at the level of individuals with substantial variation in flowering duration between flowers (Fig. 5). Our tracking algorithm was able to track individual flowers with consistently high accuracy (MOTA $\geq$ 0.981) across all series. In all cases, all flowers were tracked with more than 80% accuracy (Table 2) and two series were tracked without errors (2018 THUL B and 2018 NARS D). Naive tracking returned high MOTA scores across series, but for all three series that were not tracked perfectly, utilizing the tracking parameters increased the MOTA score.

Importantly, the tracking that returned the lowest number of mismatches did not necessarily result in a number of tracks equal to the number of flowers in a plot. For example, for the series with the most flowers, 2019 NARS C, one combination of parameters (max gap = 20, max distance = 600, running mean = 1), returned the lowest number of mismatches (31), but the number of tracks were lower than the actual number of flowers (71 vs. 85). One paramereter combination did return a number of tracks equal to number of flowers (max gap = 10, max distance 400, running mean = 10), but also increased the number of mismatches to 91. For 2019 NARS D, several combinations of parameters returned a perfect MOTA score, but still some combinations returned a number of tracks lower than the number of flowers. This demonstrates the significance of merging errors, which are not accounted for in the MOTA score. In figure 6, we exemplify the two types of errors, identity shifts and merges, in the flower data.

![Figure 5: Flowering lifespan (days) for each individual flower in each of the five image series plotted against the onset of flowering in days after 1 January (day of the year) based on manual tracking of individual flowers](../figures/OnsetVSLengthBoxWithPoints.png){width="100%"}

@@

![Figure 6: Tracking results of three flowers in the 2018 NARS A image series. A: Track output from the flower tracking algorithm for three individual flowers using two different combinations of tracking parameters. The y-axis shows the correct track IDs assigned by manual tracking. Each track is coloured by the IDs assigned by the flower tracking algorithm. The top panel (max gap = 10, running mean = 1, max distance = 0) shows a merging error of flower 15 and 28 (see B), while flower 25 has been fragmented into two separate tracks causing one mismatch (see C). The bottom panel show correct tracking output returned when using the parameters max gap = 10, running mean = 20, max distance = 50. B: Merging error between two tracks. Frame f shows the last frame containing flower 15 (blue). Two frames (5 hours) later (f+2), flower 28 (blue) has  bolted. Since max gap \> 2 and no distance threshold has been set, flower 28 is associated with the track of flower 15. Setting a max distance threshold to 500 fixed this error as illustrated in A. C: Displacement of flowers 25 and 29 between two frames, f and f+1, (7 hours apart), causes an identity shift. Setting running mean to 20 fixed this error as illustrated in A.](../figures/errorExamples.png){width="100%"}

With our filtering method, we were able to identify 28 tracks with no mismatches from the three series (2018 NARS A: 15 tracks; 2019 NARS C: 10 tracks; 2019 NYAA E: 3 tracks) with complex scenes using a fixed filtering distance value of 350 and fixed values for the tracking parameters (suppl. 2). The results of the filtering algorithm on the three series are given in Table 3 and the filtering process is visualised in figure 7.

@@

![Figure 7: The track filtering process exemplified by the 2019 NARS C series. A: The centroid points for all bounding boxes in the series, coloured by the track id from the centroid tracking algorithm. B: The track convex hull polygons calculated from the tracks. Polygons are coloured according to group and polygons with a euclidean distance below 350 between them have been grouped together. C: The results of the filtering where all tracks that were not sufficiently spatially isolated have been removed, resulting in 10 remaining tracks without tracking errors.](images/filterFigLet.png)

@@

In this paper, we have demonstrated the potential of automated tracking to monitor floral phenology of individual flowers from bolting to withing and produce new insight into floral biology. We have presented a simple distance-based tracking algorithm and, with simulated data, we have shown how tracking parameters can be used to tune the tracking algorithm to deal with specific challenges and increase tracking performance. We have tested our method on real flower and demonstrated the feasibility of tracking individual flowers in images through time. Challenges similar to the ones demonstrated through simulations existed in the real flower data and adjustment of the tracking parameters allowed for increased tracking accuracy of the real flowers.

The automatic tracking and filtering pipeline presented was able to return reliable flower tracks despite complex scenes with a high number of objects with sporadic visibility and violation of standard tracking algorithm assumptions. Even with simple centroid tracking where association is based solely on displacement between the current and previous frame, tracking performances were high for all flower series and perfect for two. Further, coupling this the user-defined tracking parameters and our filtering approach, we were able to optimize tracking performance and extract reliable tracks.

The two series for which the tracking algorithm produced perfect results were also the series with the lowest number of flowers. While an increased number of objects does not necessarily increase tracking difficulty as this depends on the degree of isolation of tracks in time and space [@beard2018], here, since the flowers are limited in space and time, the complexity of tracking is likely to increase with higher numbers of flowers as this increases the possibility of flowers overlapping or moving in and out of frame. The high complexity in the scenes with many flowers, especially 2019 NARS C with 85 flowers, coupled with the low frame rate renders it unlikely that any tracking method could produce perfect results across series. Although 2019 NARS C may be considered an extreme case for our application, it underlines the relevance of a pipeline including both tracking and filtering, but also points to the possibility of filtering image series based on an estimate of the number of flowers before automated tracking, which could be relevant in some cases.

Importantly, optimizing for high MOTA can lead to an underestimate of the number of tracks (i.e., merging of tracks occurs) (Table 2). While accepting gaps in tracks allows for temporary occlusion of a flower within a track it may also lead to merging of separate tracks. Flower tracking may be especially prone to these errors since flowers can bolt in the same area at different times during the season. The task of evaluating multiple object tracking accuracy is a challenge in itself [@luiten2020; @milan2013; @milan2016] and other tracking accuracy metrics that account for merging errors in a better way should be further studied.

The optimal settings for the three tracking parameters fell into a narrow range for the three complex series, except for the running mean for 2019 NYAA E, which were exceptionally high compared to the other series. Interestingly, the average flower duration was also higher for this series and this could be part of the explanation, as longer lifespan allow for larger gaps in the tracks. The three series with complex scenes all contain flowers at the edge of the frame that may occasionally be outside the view of the camera (see Fig. 3) and the high number of flowers increase the risk of flowers temporarily occluding others. Correct tracking of these objects requires a max gap setting that optimizes the trade-off between fragmentation and merging errors. Overall, the max gap parameter improved tracking accuracy with values of 10-20 being optimal.

The optimal range for max distance was relatively narrow (400-600) for the parameters producing the correct number of tracks with the lowest number of mismatches. The optimal value for this setting is one that is sufficiently high that it does not force individual flowers into separate tracks because of movement between frames, yet is low enough to separate flowers into individual tracks that would otherwise have been grouped together. This explains why the extreme values for this parameter was not included in the highest accuracy tracks.

It is important to note that the tracking parameters are not independent. For example, changing the setting for running mean could influence the optimal value for max distance. This underlines the relevance of searching for an optimal parameter combination based on ground truth data. As with any tracking method, accuracy can only be assessed based on ground truth data [@dendorfer2021; @luiten2020; @milan2016]. However, the fact that the optimal tracking parameters were overall similar across series strengthens the assumptions that a set of standard values can produce reasonable results for other similar image series with *Dryas* flowers. Here, a combination such as 10, 500, 10, for max gap, max distance, and running mean, respectively, would be a good starting point. When applying our method in a new setting, we recommend performing a parameter test as demonstrated here on a subset of ground truth tracks to derive the optimal set of parameters.

We demonstrated our method on manual annotations of flowers but note that the format of the input data matches the output of deep learning object detection methods [@everingham2010]. Thus, the method can be directly applied on such data, for instance as produced by the method described in @mann2022. Automatic detecting of flowers would likely introduce false negatives when the detection algorithm fails to detect an object, but here the max gap parameter may alleviate such problems. Automatic detection methods may also introduce false positives. These detections would be included in the tracking along with correct detections. Some cases of false positives, for example isolated single false positives, could be removed automatically while other cases may require manual quality control. In general, the performance of a tracking algorithm depends on the quality of input data and it is advisable to optimise detection accuracy before moving to tracking. The fact that our method produced perfect tracking results in the most simple settings (fewest flowers) could be taken into consideration when setting up future data collection schemes.

>>>>

```{r eval = FALSE, echo=FALSE}
library(knitr)
knit('ManuscriptDraft.Rmd')
rmarkdown::render("ManuscriptDraft.Rmd", output_file = "ManuscriptDraft_Resubmission.docx")
system2("open","ManuscriptDraft_Resubmission.docx")
```

```{r echo=FALSE}
# Function for creating word comments in Rmarkdown
# Use: This text contains a `r word_comment("This is the comment", "comment.")`.

@@

Keywords: Arctic, *Dryas octopetala*, *Dryas integrifolia*, ecological monitoring, floral phenology, life-history variation, tracking

@@

Article type: Research article

Word count: 8674

@@

1.  A fundamental challenge in floral biology is to understand how reproductive success is affected by the timing of flowering, the timing and frequency of pollinator visits, and the abundance and diversity of visitors. Since plant-pollinator interactions concern individual flowers, deciphering the consequences of e.g. shifts in the frequency of flower visits for plant reproductive success should ideally be based on observations of individual flowers throughout their entire flowering period. Due to the overwhelming logistical challenge of collecting such data by direct observation, it has rarely been done.

2.  Here, we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of individual flowers. Further, we present a framework for automatic tracking of flowers in time-lapse image series, evaluating tracking accuracy, and identifying reliable tracks.

@@

The Arctic is characterized by short growing seasons and for the individual plant, timing of flowering is of critical importance. Flowering too early increases the risk of frost damage to flowers, while flowering too late limits the time for seed maturation and dispersal before the end of the growing season [@pardee2019; @inouye2008; @iler2021]. Further, for insect-pollinated plants, flowering must be synchronous with pollinator activity for successful reproduction. Both flower and insect phenology is partly driven by abiotic cues such as snowmelt and temperature and can thus be sensitive to climate change. Changes in floral phenology as a consequence of climate change may indirectly affect the abundance of flower-dependent insect species [@hoye2013] and these effects can be stronger than the direct effects of climate change [@ogilvie2017]. However, the relation between floral phenology and reproductive success is not well understood. For example, @bolmgren2015 found that for an insect-pollinated plant species, early flowering was associated with increased fruit-set but late flowering was associated with higher seed mass. If we are to understand the intricate dynamics of plant-pollinator interactions, tools that aid efficient and comprehensive data collection are essential.

@@

Cameras have become an ubiquitous tool for non-invasive and efficient monitoring of natural processes [@howe2017; @glover2019], but the implementation of image-based data collection has outpaced the development of methods for processing and analysing the big data output [@weinstein2018; @farley2018]. Gradually, computer vision and deep learning tools are being developed and applied for automatic extraction of relevant data from images [@willi2019; @hoye2021]. Indeed, @mann2022 demonstrated that automatic image-based monitoring of flowering phenology coupled with deep learning can facilitate collection of phenology data for specific species at very high temporal resolution. While such automated detection of flowers can produce detailed information about floral biology, phenological responses at the level of individual flowers may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be directly explored on the basis of individual level floral phenology data and such information is critical for investigating the association between reproductive success and timing of flowering and/or flower longevity. Individual level data would also aid the understanding of whether flower visitation rates and/or seed set depends on the timing of flowering for the individual flower. Thus, automated recording of individual floral phenology by tracking the presence of individual flowers through time would be ideal and the feasibility of this as a step subsequent of flower detection was explored in this study.

@@

Many multiple object tracking (MOT) methods focus on tracking of humans [@luo2021], and are not necessarily directly transferable to other use cases. Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], rely on Kalmann filtering to predict the future location of an object based on its previous velocity [@kalmann]. However, Kalman filtering assumes linear object motion and does not handle abrupt movement well. Flowers are constrained in movement by their stalk and violate this assumption. Other tracking approaches, such as DeepSORT [@deepsort2017], use tracking-by-detection in which convolutional neural networks (CNNs) are applied to continuously distinguish individuals through time, thus aiding the tracking algorithm. Individual flowers, however, are unlikely to be clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame, which prevents this approach. Finally, some algorithms require manual initiation of a fixed number of tracks in the first frame and do not allow objects to enter or disappear during tracking [@hu2012]. In full-season monitoring data, flowers should continuously be tracked as they bolt and wilt throughout the season (fig. 1). Additionally, a flower tracking method must handle sporadic visibility of flowers because of occlusion or temporary movement outside of frame view. Clearly, individual flower tracking is a challenging task.

@@

In MOT, tracking accuracy is quantified by comparing tracking output with a ground truth [@dendorfer2021; @luiten2020; @milan2013] and two fundamental errors can occur. First, a single trajectory may be split into several tracks, an error known as fragmentation. This can happen when an object is temporarily occluded or have disappeared from view and a new track is erroneously initiated when it reappears, or when identity shifts occur between tracks, for example when they switch positions between frames. Fragmentation is especially prone in settings with high densities of objects. Second, merging occurs when several trajectories are assigned the same track identity. This can occur when a track is not correctly terminated at the end of a trajectory but rather continues on the trajectory of another object. Here, importantly, the two individual trajectories can be tracked without internal errors (i.e. without mismatches).

@@

![Figure 2: Simulated flower detection data to demonstrate the tracking algorithm and the effect of the three tracking parameters, max distance, max gap, and running mean (refer to Automatic flower tracking section for definitions), as well as the significance of frame rate for reliable tracking. A: 10 points per flower, rF (radius of circle delimiting flower points) = 200, rM (radius of circles delimiting distance between consecutive points) = 100. Arrow shows the last point of flower a and the first point of flower b. B: 10 points per flower, rF = 200, rM = 100. C: 500 points per flower, rF = 200, rM = 200. D: 10,000 points per flower, rF = 100, rM = 1.](../figures/simulatedData_v2.png)

@@

@@

![Table 1: Number of images and number of individual flowers in each of the five time-lapse image series.](../figures/seriesOverview.png)

![Figure 3: Examples of annotated images for 2019 NARS C. A: Onset of flowering at DOY 145. B: Peak flowering at day of year (DOY) 151 (52 annotated flowers). C: 6 days after peak flowering (DOY 157). Flower bounding boxes are colour-coded by flower ID. D: Centroids of all flower annotations in the series, coloured by flower ID, showing how flowers may overlap in space.](../figures/AnnotationExamples_v2_5.png)

@@

We present a framework for tracking, filtering, and evaluating tracking of objects in time-lapse image series. The tracking algorithm (suppl. 1) is based on centroid tracking [@rahman2020; @bjerge2021]. That is, objects with minimum displacement between frames are associated. Centroids are derived from bounding boxes around objects stemming from manual annotation or automatic detection of objects. Thus, the input data for the tracking algorithm are bounding box coordinates (top left corner and bottom right corner) for each object [@everingham2010]. The algorithm has a set of user-adjusted parameters that can optimize tracking accuracy. The parameters are particularly relevant for tracking of objects that are constrained to a specific area such as plants.

Several issues complicate the task of tracking individual flowers through time-lapse image series. First, as flowers move, they may temporarily occlude the view of other flowers and flowers close to the edge of the frame may move in and out of view between images. This can cause a track to be lost and fragmentation to occur. However, if a flower reappears shortly after having disappeared from view, it is a reasonable assumption that it is the same individual. The parameter **max gap** sets the number of frames a flower can be absent before reappearing without triggering track termination and initiation of a new track. Concurrently, the parameter can prevent track termination owing to false negatives.

Second, a standard centroid tracking algorithm will associate a point with a track based on minimum distance only, disregarding the absolute distance between centroids, which may lead to tracking errors. However, as flowers are constrained in their displacement, we can assume a maximum distance between points and force the initiation of tracks for points that exceed this threshold. We set this threshold with the parameter **max distance,** measured as the euclidean distance in number of pixels between objects in neighboring frames.

Third, reliable tracking based on distance requires that movement of individuals are captured at sufficient frequency to prevent identity shifts. Stand-alone image-based monitoring of the relatively slow process of flower development through full growing seasons (several months) is likely to be done with low frame rate, limited by power consumption and storage capacity both on and off the camera. Here, we used several images per day and while this temporal resolution captures the phenology of the flowers at very high detail, they fail to capture the full movement of flowers. For example, as the wind shifts, the flower heads change direction, which in the image series happens instantaneously (i.e., between two consecutive frames). With a high number of flowers in close vicinity to each other, this leads to a substantial risk of identity shifts between flowers (Fig. 2C). Additionally, as the flowers are constrained by their stalk, there is a limit to the distance they can move, which prevents prediction of future location of an individual based on its previous trajectory. Here, we use the tracking parameter **running mean**, with which we can base tracking on the distance between a point in the current frame and the running mean of the positions of a number of previous points in a track, facilitating a reduction of the weighting on movement between individual frames. Setting the parameter to 1 means that tracking is based on the distance to the centroid in the previous frame only.

@@

## Flowers

Manual tracking of individual flowers derived from image-based plot monitoring enabled detailed information on floral phenology at the level of individuals with substantial variation in flowering duration between flowers (Fig. 4). Our tracking algorithm was able to track individual flowers with consistently high accuracy (MOTA $\geq$ 0.981) across all series. In all cases, all flowers were tracked with more than 80% accuracy (Table 2) and two series were tracked without errors (2018 THUL B and 2018 NARS D). Naive tracking returned high MOTA scores across series, but for all three series that were not tracked perfectly, utilizing the tracking parameters increased the MOTA score.

Importantly, the tracking that returned the lowest number of mismatches did not necessarily result in a number of tracks equal to the number of flowers in a plot. For example, for the series with the most flowers, 2019 NARS C, one combination of parameters (max gap = 20, max distance = 600, running mean = 1), returned the lowest number of mismatches (31), but the number of tracks were lower than the actual number of flowers (71 vs. 85). One paramereter combination did return a number of tracks equal to number of flowers (max gap = 10, max distance 400, running mean = 10), but also increased the number of mismatches to 91. For 2019 NARS D, several combinations of parameters returned a perfect MOTA score, but still some combinations returned a number of tracks lower than the number of flowers. This demonstrates the significance of merging errors, which are not accounted for in the MOTA score. In figure 5, we exemplify the two types of errors, identity shifts and merges, in the flower data.

![Figure 4: Flowering lifespan (days) for each individual flower in each of the five image series plotted against the onset of flowering in days after 1 January (day of the year) based on manual tracking of individual flowers](../figures/OnsetVSLengthBoxWithPoints.png){width="100%"}

@@

![Figure 5: Tracking results of three flowers in the 2018 NARS A image series. A: Track output from the flower tracking algorithm for three individual flowers using two different combinations of tracking parameters. The y-axis shows the correct track IDs , that is the ground truth IDs assigned by manual tracking. Each track is coloured by the IDs assigned by the flower tracking algorithm. For correct algorithmic tracking, a whole groun truth track should have one unique color. The top panel (max gap = 10, running mean = 1, max distance = 0) shows a merging error of flower 15 and 28 (see B), while flower 25 has been fragmented into two separate tracks causing one mismatch (see C). The bottom panel show correct tracking output returned when using the parameters max gap = 10, running mean = 20, max distance = 50. B: Merging error between two tracks. Frame f shows the last frame containing flower 15 (blue). Two frames (5 hours) later (f+2), flower 28 (blue) has bolted. Since max gap \> 2 and no distance threshold has been set, flower 28 is associated with the track of flower 15. Setting a max distance threshold to 500 fixed this error as illustrated in A. C: Displacement of flowers 25 and 29 between two frames, f and f+1, (7 hours apart), causes an identity shift. Setting running mean to 20 fixed this error as illustrated in A.](../figures/errorExamples.png){width="100%"}

With our filtering method, we were able to identify 28 tracks with no mismatches from the three series (2018 NARS A: 15 tracks; 2019 NARS C: 10 tracks; 2019 NYAA E: 3 tracks) with complex scenes using a fixed filtering distance value of 350 and fixed values for the tracking parameters (suppl. 2). The results of the filtering algorithm on the three series are given in Table 3 and the filtering process is visualised in figure 6.

@@

![Figure 6: The track filtering process exemplified by the 2019 NARS C series. A: The centroid points for all bounding boxes in the series, coloured by the track id from the centroid tracking algorithm. B: The track convex hull polygons calculated from the tracks. Polygons are coloured according to group and polygons with a euclidean distance below 350 between them have been grouped together. C: The results of the filtering where all tracks that were not sufficiently spatially isolated have been removed, resulting in 10 remaining tracks without tracking errors.](images/filterFigLet.png)

@@

In this paper, we have demonstrated the potential of automated tracking to monitor floral phenology of individual flowers from bolting to withing to produce new insight into floral biology. We have presented a simple distance-based tracking algorithm and, with simulated data, we have shown how a set of tracking parameters can be used to tune the tracking algorithm to deal with specific challenges and increase tracking performance. We have tested our method on real flower data and demonstrated the feasibility of tracking individual flowers in images through time. Challenges similar to the ones demonstrated through simulations existed in the real flower data and while even simple centroid tracking resulted in high accuracy across all flower series and perfect accuracy for two, adjustment of the tracking parameters allowed for increased tracking accuracy of the flowers. Further, by coupling tracking with our suggested filtering approach, we were able to extract reliable tracks in complex scenes. Overall, the automatic tracking and filtering pipeline presented was able to return reliable flower tracks despite complex scenes with a high number of objects with sporadic visibility and violation of standard tracking algorithm assumptions.

The two series for which the tracking algorithm produced perfect results were also the series with the lowest number of flowers. While an increased number of objects does not necessarily increase tracking difficulty as this depends on the degree of isolation of tracks in time and space [@beard2018], here, since the flowers are limited in space and time, the complexity of tracking is likely to increase with higher numbers of flowers as this increases the possibility of flowers overlapping or moving in and out of frame. The high complexity in the scenes with many flowers, especially 2019 NARS C with 85 flowers, coupled with the low frame rate renders it unlikely that any tracking method could produce perfect results across series. Although 2019 NARS C may be considered an extreme case for our application, it underlines the relevance of a pipeline including both tracking and filtering, but also points to the possibility of filtering image series based on an estimate of the number of flowers before automated tracking, which could be relevant in some cases. The fact that our method produced perfect tracking results in the most simple settings (fewest flowers) could be taken into consideration when setting up future data collection schemes.

Importantly, optimizing for high MOTA can lead to an underestimate of the number of tracks (i.e., merging of tracks occurs) (Table 2). While accepting gaps in tracks allows for temporary occlusion of a flower within a track it may also lead to merging of separate tracks. Flower tracking may be especially prone to these errors since flowers can bolt in the same area at different times during the season. The task of evaluating MOT accuracy is a challenge in itself [@luiten2020; @milan2013] and other tracking accuracy metrics that account for merging errors in a better way should be further studied.

The optimal settings for the three tracking parameters fell into a narrow range for the three complex series, except for the running mean for 2019 NYAA E, which were exceptionally high compared to the other series. Interestingly, the average flower duration was also higher for this series and this could be part of the explanation, as longer lifespan allow for larger gaps in the tracks. The three series with complex scenes all contain flowers at the edge of the frame that may occasionally be outside the view of the camera (see fig. 3) and the high number of flowers increase the risk of flowers temporarily occluding others. Correct tracking of these objects requires a max gap setting that optimizes the trade-off between fragmentation and merging errors. Overall, the max gap parameter improved tracking accuracy with values of 10-20 being optimal.

The optimal range for max distance was relatively narrow (400-600) for the parameters producing the correct number of tracks with the lowest number of mismatches. The optimal value for this setting is one that is sufficiently high that it does not force individual flowers into separate tracks because of movement between frames, yet is low enough to separate flowers into individual tracks that would otherwise have been grouped together, which explains why the extreme values for this parameter was not included in the highest accuracy tracks.

The best values for running mean were either 1 or 10, the two lowest possible values. The parameter may have had an effect in regards to identifying the correct number of tracks, but this is not immediately clear from the results, since it always differed in combination with others. In future studies it may be relevant to include values between 0 and 10.

It is important to note that the tracking parameters are not independent. For example, changing the setting for running mean could influence the optimal value for max distance. This underlines the relevance of searching for an optimal parameter combination based on ground truth data. However, the fact that the optimal tracking parameters were overall similar across series strengthens the assumptions that a set of standard values can produce reasonable results for other similar image series with *Dryas* flowers. Here, a combination such as 10, 500, 10, for max gap, max distance, and running mean, respectively, would be a good starting point. When applying our method in a new setting, we recommend performing a parameter test as demonstrated on a subset of ground truth tracks to derive the optimal set of parameters.

We demonstrated our method on manual annotations of flowers but note that the format of the input data matches the output of deep learning object detection methods [@everingham2010]. Thus, the method can be directly applied on such data, for instance as produced by the method described in @mann2022. Automatic detecting of flowers would likely introduce false negatives when the detection algorithm fails to detect an object, but here the max gap parameter may alleviate such problems. Automatic detection methods may also introduce false positives. These detections would be included in the tracking along with correct detections. Some cases of false positives, for example isolated single false positives, could be filtered automatically while other cases may require manual quality control. In general, the performance of a tracking algorithm depends on the quality of input data and it is advisable to optimise detection accuracy before moving to tracking.
